``Colloidal rain'' as a novel route to crystallization via viscoelastic phase separation

### 2015-09-29 science/crystal_gel_letter2.tex
Updated some references. Minor changes to the text. Written first draft of the cover letter.

\author{Hideyo Tsurusawa$1$, John Russo$^1$, Mathieu Leocmach$^{2}$ and Hajime Tanaka$^{1,\ast}$ 
\\
\normalsize{$^1$Institute of Industrial Science, University of Tokyo,}\\
\normalsize{4-6-1 Komaba, Meguro-ku, Tokyo 153-8505, Japan}\\
\normalsize{$^2$Institut Lumière Matière, CNRS UMR 5306, Université Claude Bernard Lyon 1, }\\
\normalsize{Université de Lyon, Lyon, 69622 Villeurbanne Cedex, France}\\
\normalsize{$^\ast$Corresponding author. E-mail: tanaka@iis.u-tokyo.ac.jp}}

@@

\maketitle
\vspace{-1cm}
\begin{sciabstract}
%Colloidal gels form when network-forming spinodal phase separation is interrupted by the glass transition~\cite{lu2008gelation}. It has been shown that crystallization may equivalently interrupt the phase separation by freezing the dense phase. However the particle-level details of this process have not yet been confirmed experimentally. 
Here we show by confocal microscopy measurements that crystallization leads to a network structure different from the original phase separation pattern. Crystals nucleate inside the liquid network, and grow past it by direct condensation of the gas phase on their surface, driving evaporation of nearby liquid. This process represents the colloidal analogue of the Bergeron process, i.e., ice crystals formation that occurs in mixed phase clouds and is at the origin of rain. Our \emph{colloidal cold rain} gives us full experimental access to the kinetic pathway of this phenomenon important for climate science. We argue that states similar to colloidal cold rain can form as the result of the gas-liquid phase separation of crystallizable components, such as monoatomic and molecular systems, whose dynamics is slow enough to induce viscoelastic phase separation, but fast enough to prevent immediate vitrification.
\end{sciabstract}

@@

Gels and glasses are two important non-ergodic disordered states, which are very important 
in our life and nature~\cite{anderson2002insights,lekkerkerker2011colloids}.
The dynamical arrest of colloidal gels and glasses is believed to originate from a dynamical transition to the same non-ergodic disordered glass state, i.e., vitrification. The only difference comes from the fact that the former is formed by densification induced by phase separation and thus is spatially heterogeneous, whereas the latter is formed by uniform densification and is thus spatially homogeneous.
Gels are distinct from glasses because of their heterogeneous network structure, which gives them properties intermediate between
those of a solid and a fluid phase, and allows the coexistence of two properties usually incompatible with each other, 
elasticity and fluidity. 
The elasticity of gels originates from the percolated network structure and the resulting space-spanning connectivity. 
At the same time a fluid component can easily flow through the elastic network.
There are two types of gels with such characteristics: one is a network structure stabilized by stable crosslinks, and the other is a phase-separated structure frozen by dynamical arrest due to vitrification.
Many chemical and physical gels belong to the former category, in which the size of network pores is rather small and 
the fluid transport is not so efficient. On the other hand, gels made of spherical colloids and proteins 
with attractive interactions belong to the latter category. 
This type of gels are formed by viscoelastic phase separation \cite{tanaka1999colloid,tanaka2000viscoelastic} into a dense liquid phase with slow dynamics and a dilute gas phase 
with fast dynamics. This difference in the viscoelastic properties between the two phases allows the system to form a space-spanning 
network structure of the liquid phase, even if it is the minority phase. During phase separation, the density of the liquid phase 
increases towards the glass-transition density, leading to slow glassy dynamics.
At the glass transition point, the percolated network structure is dynamically stabilised by vitrification 
of the dense liquid phase~\cite{pusey1993dynamics,ilett1995phase,verhaegh1997transient,tanaka1999colloid,foffi2002,zaccarelli2007,lu2008gelation,zaccarelli2008gelation,testard2011}.  
More precisely, the dynamical stabilization is due to percolation of locally favored structures, which are locally stable non-crystalline structures \cite{royall2008g}.
% Since ``glass'' is the term used for bulk, this may be a more appropriate way to express the mechanism of dynamical arrest. 
% This type of gels are made by phase separation. Thus, the structure is intrinsically heterogeneous, and the network pore size is usually 
% much larger than gels made of stable crosslinks, resulting in a high fluid transport ability.

@@

In this Letter, we show that, for low polymer concentration, the system can undergo a novel crystallization pathway that leads to the formation
of crystalline droplets. These droplets originate from crystal seeds that form inside the liquid branches of the spinodal aggregate, but then grow by direct condensation from the gas phase. The process is analogous to the Bergeron process of ice formation in mixed phase clouds, where ice droplets grow at the
expense of the supercooled liquid droplets due to their lower saturated vapor pressure. 
This is speculated to be the primary mechanism for the formation of raindrops. 
In our colloidal system, we then call this process ``colloidal cold rain''
formation. Due to the difference in the nucleation and growth processes, the droplets display different scaling laws depending on their
size. For small nuclei, the droplets have the same fractal dimension of the embedding gel network. But during the growth process, droplets
grow with an integer fractal dimension due to direct condensation of gas particles. This is in contrast to the standard crystal-gel scenario, where crystals nucleate and grow inside the dense branches of the gel, becoming dynamically arrested. We also observe that increasing the strength of the interparticle attraction
(by increasing the polymer concentration), instead of forming colloidal cold rain, the system undergoes the classical gelation process, in which
phase separation is interrupted by vitrification~\cite{verhaegh1997transient,tanaka1999colloid,Royall2007,lu2008gelation}.

@@

To confirm the formation of crystalline seeds we adopt bond orientational analysis to detect which particles are in crystalline environments~\cite{russo2012microscopic}. With this analysis we confirm that crystallization events start occurring for the state point $\phi=0.25$ and $c_p=0.38$, which has the lowest
polymer concentration of all state points. This is shown in Fig.~\ref{fig:crystals}A, where the black line indicates the average size of the crystallites,
while the red line the average number of crystallites as a function of time. The number of crystallites first rapidly increases as nucleation events
start occurring inside the liquid branches, but eventually starts decreasing as the different crystallites grow and merge with each other.
%The average size of the crystals shows an interesting linear growth regime for $100<t<600$. This linear growth regime goes past the time
%of coalescence of nuclei, so a first speculation is that it could be due to the compensation of the following two effects:
%\begin{itemize}
% \item diffusive growth of the nuclei, which scales like $t^{1/2}$
% \item coalescence growth, which should scale faster than $t$, maybe $t^2?$
%\end{itemize}
All other state points show only negligible signs of crystallization. In particular, for state points with $\phi=0.25$, increasing
polymer concentration drastically reduces the amount of crystals. This is in agreement with the
idea of enhanced crystallization rates near metastable critical points~\cite{ten1997enhancement,olmsted1998spinodal} and the
results of Refs.~\cite{fortini2008crystallization,perez2011pathways},
which speculated two different arrest mechanisms: crystallization at low polymer concentration, and dynamic arrest at high polymer concentrations.

While nucleation always occurs inside the liquid branches of the phase-separating network, we suggest here the presence of a new growth channel
which was neither considered nor observed before. In the following we will show that the growth mechanism of the crystal is not only due to filling of the spinodal liquid network, as was observed in Refs.~\cite{fortini2008crystallization,perez2011pathways} for Brownian Dynamics simulations, but also involves direct condensation of the gas phase
on the crystalline seeds.
To investigate this process, we plot in Fig.~\ref{fig:crystals}C the gyration radius of individual crystalline nuclei for the state point
$\phi=0.25$ and $c_p=0.38$. The results show that the crystal growth follows two different scaling laws: at small crystalline sizes it scales with
the dimension of random percolation $D=2.53$, while at large sizes it scales as $D=3$ of compact crystal growth. This demonstrates that, while
small crystalline nuclei are nucleated inside the phase-separated liquid branches, once they reach the transverse size of the liquid branch, the growth
follows a volume expansion with the formation of isotropic droplets (colloidal cold rain). The same scaling law was suggested in the analysis of the
structure factors, Fig.~\ref{fig:phase_separation}C, but it is shown directly by the analysis of the gyration radius.

@@

The process we observed is akin to the Bergeron process which is the primary mechanism for the formation of cold rain drops in clouds~\cite{glickman2000glossary}.
In clouds there is a mixture of ice crystals and supercooled water. The vapor phase is in coexistence with the liquid phase, but is supersaturated
with respect to the ice crystals. This causes the water droplets to evaporate and sublimate directly on the ice crystals. Our system can then
be regarded as a colloidal analogue for this important process, which can now be studied at the single-particle level.

>>>>

\author{Hideyo Tsurusawa$^1$, John Russo$^1$, Mathieu Leocmach$^{2}$ and Hajime Tanaka$^{1,\ast}$ 
\\
\normalsize{$^1$Institute of Industrial Science, University of Tokyo,}\\
\normalsize{4-6-1 Komaba, Meguro-ku, Tokyo 153-8505, Japan}\\
\normalsize{$^2$Institut Lumière Matière, CNRS UMR 5306, Université Claude Bernard Lyon 1, }\\
\normalsize{Université de Lyon, Lyon, 69622 Villeurbanne Cedex, France}\\
\normalsize{$^\ast$Corresponding author. E-mail: tanaka@iis.u-tokyo.ac.jp}}

@@

\maketitle
\vspace{-1cm}
\begin{sciabstract}
%Colloidal gels form when network-forming spinodal phase separation is interrupted by the glass transition~\cite{lu2008gelation}. It has been shown that crystallization may equivalently interrupt the phase separation by freezing the dense phase. However the particle-level details of this process have not yet been confirmed experimentally. 
Here we show by confocal microscopy measurements that crystallization leads to a network structure different from the original phase separation pattern. Crystals nucleate inside the liquid network, and grow past it by direct condensation of the gas phase on their surface, driving evaporation of nearby liquid. This process represents the colloidal analogue of the Bergeron process, i.e., ice crystals formation that occurs in mixed phase clouds and is at the origin of rain. Our \emph{colloidal rain} gives us full experimental access to the kinetic pathway of this phenomenon important for climate science. We argue that states similar to colloidal rain can form as the result of the gas-liquid phase separation of crystallizable components, such as monoatomic and molecular systems, whose dynamics is slow enough to induce viscoelastic phase separation, but fast enough to prevent immediate vitrification.
\end{sciabstract}

@@

Gels and glasses are two important non-ergodic disordered states, which are very important 
in our life and nature~\cite{anderson2002insights,lekkerkerker2011colloids}.
The dynamical arrest of colloidal gels and glasses is believed to originate from a dynamical transition to the same non-ergodic disordered glass state, i.e., vitrification. The only difference comes from the fact that the former is formed by densification induced by phase separation and thus is spatially heterogeneous, whereas the latter is formed by uniform densification and is thus spatially homogeneous.
Gels are distinct from glasses because of their heterogeneous network structure, which gives them properties intermediate between
those of a solid and a fluid phase, and allows the coexistence of two properties usually incompatible with each other, 
elasticity and fluidity. 
The elasticity of gels originates from the percolated network structure and the resulting space-spanning connectivity. 
At the same time a fluid component can easily flow through the elastic network.
There are two types of gels with such characteristics: one is a network structure stabilized by stable crosslinks, and the other is a phase-separated structure frozen by dynamical arrest due to vitrification.
Many chemical and physical gels belong to the former category, in which the size of network pores is rather small and 
the fluid transport is not so efficient. On the other hand, gels made of spherical colloids and proteins 
with attractive interactions belong to the latter category. 
This type of gels are formed by viscoelastic phase separation \cite{tanaka1999colloid,tanaka2000viscoelastic} into a dense liquid phase with slow dynamics and a dilute gas phase 
with fast dynamics. This difference in the viscoelastic properties between the two phases allows the system to form a space-spanning 
network structure of the liquid phase, even if it is the minority phase. The network is fractal and its dynamic structure factor shows a stretched-exponential decay to a finite
plateau~\cite{krall1998internal,solomon2001dynamic,romer2000sol}. During phase separation, the density of the liquid phase 
increases towards the glass-transition density, leading to slow glassy dynamics.
At the glass transition point, the percolated network structure is dynamically stabilised by vitrification 
of the dense liquid phase~\cite{pusey1993dynamics,ilett1995phase,verhaegh1997transient,tanaka1999colloid,foffi2002,buzzaccaro2007sticky,zaccarelli2007,lu2008gelation,zaccarelli2008gelation,testard2011}.  
More precisely, the dynamical stabilization is due to percolation of locally favored structures, which are locally stable non-crystalline structures \cite{royall2008g}.
% Since ``glass'' is the term used for bulk, this may be a more appropriate way to express the mechanism of dynamical arrest. 
% This type of gels are made by phase separation. Thus, the structure is intrinsically heterogeneous, and the network pore size is usually 
% much larger than gels made of stable crosslinks, resulting in a high fluid transport ability.

@@

In this Letter, we show that, for low polymer concentration, the system can undergo a novel crystallization pathway that leads to the formation
of crystalline droplets. These droplets originate from crystal seeds that form inside the liquid branches of the spinodal aggregate, but then grow by direct condensation from the gas phase. The process is analogous to the Bergeron process of ice formation in mixed phase clouds, where ice droplets grow at the
expense of the supercooled liquid droplets due to their lower saturated vapor pressure. 
This is speculated to be the primary mechanism for the formation of raindrops. 
In our colloidal system, we then call this process ``colloidal rain''
formation. Due to the difference in the nucleation and growth processes, the droplets display different scaling laws depending on their
size. For small nuclei, the droplets have the same fractal dimension of the embedding gel network. But during the growth process, droplets
grow with an integer fractal dimension due to direct condensation of gas particles. This is in contrast to the standard crystal-gel scenario, where crystals nucleate and grow inside the dense branches of the gel, becoming dynamically arrested. We also observe that increasing the strength of the interparticle attraction
(by increasing the polymer concentration), instead of forming colloidal rain, the system undergoes the classical gelation process, in which
phase separation is interrupted by vitrification~\cite{verhaegh1997transient,tanaka1999colloid,Royall2007,lu2008gelation}.

@@

To confirm the formation of crystalline seeds we adopt bond orientational analysis to detect which particles are in crystalline environments~\cite{russo2013interplay}. With this analysis we confirm that crystallization events start occurring for the state point $\phi=0.25$ and $c_p=0.38$, which has the lowest
polymer concentration of all state points. This is shown in Fig.~\ref{fig:crystals}A, where the black line indicates the average size of the crystallites,
while the red line the average number of crystallites as a function of time. The number of crystallites first rapidly increases as nucleation events
start occurring inside the liquid branches, but eventually starts decreasing as the different crystallites grow and merge with each other.
%The average size of the crystals shows an interesting linear growth regime for $100<t<600$. This linear growth regime goes past the time
%of coalescence of nuclei, so a first speculation is that it could be due to the compensation of the following two effects:
%\begin{itemize}
% \item diffusive growth of the nuclei, which scales like $t^{1/2}$
% \item coalescence growth, which should scale faster than $t$, maybe $t^2?$
%\end{itemize}
All other state points show only negligible signs of crystallization. In particular, for state points with $\phi=0.25$, increasing
polymer concentration drastically reduces the amount of crystals. This is in agreement with the
idea of enhanced crystallization rates near metastable critical points~\cite{ten1997enhancement,olmsted1998spinodal} and the
results of Refs.~\cite{fortini2008crystallization,perez2011pathways},
which speculated two different arrest mechanisms: crystallization at low polymer concentration, and dynamic arrest at high polymer concentrations.

While nucleation always occurs inside the liquid branches of the phase-separating network, we suggest here the presence of a new growth channel
which was neither considered nor observed before. In the following we will show that the growth mechanism of the crystal is not only due to filling of the spinodal liquid network, as was observed in Refs.~\cite{fortini2008crystallization,perez2011pathways} for Brownian Dynamics simulations, but also involves direct condensation of the gas phase
on the crystalline seeds.
To investigate this process, we plot in Fig.~\ref{fig:crystals}C the gyration radius of individual crystalline nuclei for the state point
$\phi=0.25$ and $c_p=0.38$. The results show that the crystal growth follows two different scaling laws: at small crystalline sizes it scales with
the dimension of random percolation $D=2.53$, while at large sizes it scales as $D=3$ of compact crystal growth. This demonstrates that, while
small crystalline nuclei are nucleated inside the phase-separated liquid branches, once they reach the transverse size of the liquid branch, the growth
follows a volume expansion with the formation of isotropic droplets (colloidal rain). The same scaling law was suggested in the analysis of the
structure factors, Fig.~\ref{fig:phase_separation}C, but it is shown directly by the analysis of the gyration radius.

@@

The process we observed is akin to the Bergeron process which is the primary mechanism for the formation of rain drops in clouds~\cite{glickman2000glossary,morrison2012resilience,}.
In clouds there is a mixture of ice crystals and supercooled water. The vapor phase is in coexistence with the liquid phase, but is supersaturated
with respect to the ice crystals. This causes the water droplets to evaporate and sublimate directly on the ice crystals. Our system can then
be regarded as a colloidal analogue for this important process, which can now be studied at the single-particle level.
